Performance Evaluation of Big Data Processing Strategies for Neuroimaging

### 2018-12-03 paper.tex
added spark section

Big Data processing engines leverage data locality, 
in-memory computing, and lazy evaluation to reduce the impact of data 
manipulation on the performance of applications. \todo{We need to 
define data locality, in-memory computing and lazy evaluation somewhere 
in the intro.} From early MapReduce 
implementations~\cite{dean2008mapreduce} to Apache 
Spark~\cite{zaharia2016apache} and now Dask~\cite{rocklin2015dask}, 
these frameworks have become 
mainstream tools for data analytics. Meanwhile, several scientific 
domains including bioinformatics, physics or astronomy, have entered 
the Big Data era due to increasing data volumes and variety. 
However, the adoption of Big Data engines for scientific data analysis 
remains limited, perhaps due to the widespread availability of 
scientific processing engines such as Pegasus~\cite{deelman2005pegasus}, 
Taverna~\cite{oinn2004taverna}, and many others.

Scientific applications slightly differ from the typical Big Data use 
cases, which might explain the remaining gap between Big Data and 
scientific engines. While Big Data applications mostly target text 
processing (Web search, frequent pattern mining, recommender systems, 
etc~\cite{leskovec2014mining}), implemented in consistent software libraries, and 
running on clouds or dedicated commodity clusters with locality-aware 
file systems such as HDFS~\cite{shvachko2010hadoop}, scientific applications often 
involve binary data such as images and signals, processed by 
command-line tools using a mix of programming languages (C, Fortran, 
Python, shell scripts) and deployed on large, shared clusters where 
data is fetched from data nodes to compute nodes using shared file 
systems such as Lustre~\cite{schwan2003lustre}. Such differences in 
applications and infrastructures have important consequences. To 
mention only one, in-memory computing requires instrumentation to be 
applied to command-line tools. \todo{Also mention that performance is 
only one of the concerns of scientific processing engines, which might
explain why Big Data engines are overlooked.}

@@

Answers to these questions have important implications. 
In~\cite{mehta2017comparative}, a comparative study of Dask, Spark, 
TensorFlow, MyriaDB, and SciDB on neuroinformatics use-cases is 
presented. It  concludes that these systems need to be extended to 
better address scientific code integration, data partitioning, data 
formats and system tuning. The work in~\cite{samba} also extended Spark 
with a better provenance system for scientific applications. We argue 
that such efforts should only be conducted if substantial performance 
improvements are expected from in-memory computing, lazy 
evaluation\todo{are we talking about lazy evaluation at all?} or data 
locality. On the other hand, scientific data processing engines are 
still being developed, and it is legitimate to wonder whether these 
projects should just be migrated to Spark, Dask, or other Big Data 
engines.

@@

>>>>

Big Data processing engines specialize in minimizing data transfers, through 
use of concepts such as data locality, in-memory computing and lazy evaluation,
to speed up the processing time of such large datasets.
Data locality was popularized by the 
MapReduce~\cite{dean2008mapreduce} framework. Rather than performing costly 
data transfers across the network, tasks were scheduled to the nodes where data
was located. Apache Spark~\cite{zaharia2016apache} succeeded MapReduce with its
incorporation of in-memory computing and lazy-evaluation which brought 
significant speed-ups in Big Data applications. In-memory computing
enabled data to be cached in memory whenever possible, reducing more costly
disk I/Os between pipeline steps. Introducing lazy
evaluation meant that results were only
computed only when necessary, enabling further performace optimizations. 
Frameworks such as
MapReduce and Spark have since become mainstream tools for data analytics, 
although many others, such as language-specific 
Dask~\cite{rocklin2015dask}, are emerging. 
Meanwhile, several scientific 
domains including bioinformatics, physics or astronomy, have entered 
the Big Data era due to increasing data volumes and variety. 
However, the adoption of Big Data engines for scientific data analysis 
remains limited, perhaps due to the widespread availability of 
scientific processing engines such as Pegasus~\cite{deelman2005pegasus}, 
Taverna~\cite{oinn2004taverna} and the necessary adaptations that would need to
be made to Big Data processing engines for scientific computing.

Scientific applications differ from the typical Big Data use 
cases, which might explain the remaining gap between Big Data and 
scientific engines. While Big Data applications mostly target text 
processing (e.g. Web search, frequent pattern mining, recommender 
systems~\cite{leskovec2014mining}), implemented in consistent software 
libraries and running on clouds or dedicated commodity clusters with locality-aware 
file systems such as HDFS~\cite{shvachko2010hadoop}, scientific applications 
often involve binary data such as images and signals, processed by 
command-line tools using a mix of programming languages (C, Fortran, 
Python, shell scripts) and are deployed on large, shared clusters where 
data is fetched from data nodes to compute nodes using centralized shared file 
systems such as Lustre~\cite{schwan2003lustre}. Such differences in 
applications and infrastructures have important consequences. To 
mention only one, in-memory computing requires instrumentation to be 
applied to command-line tools. Moreover, performance is only one of the important
aspects to scientific computing. Other important aspects include reproducibility 
and provenance tracking. While these are minimally addressed by Big Data
processing engines, it is currently insufficient for scientific computing. 
Incorporation of provenance techniques into frameworks such are Spark are
currently being investigated~\cite{samba}.

@@

Answers to these questions have important implications. 
In~\cite{mehta2017comparative}, a comparative study of Dask, Spark, 
TensorFlow, MyriaDB, and SciDB on neuroinformatics use-cases is 
presented. It concludes that these systems need to be extended to 
better address scientific code integration, data partitioning, data 
formats and system tuning. We argue 
that such efforts should only be conducted if substantial performance 
improvements are expected from in-memory computing, lazy 
evaluation\todo{are we talking about lazy evaluation at all?} or data 
locality. On the other hand, scientific data processing engines are 
still being developed, and it is legitimate to wonder whether these 
projects should just be migrated to Spark, Dask, or other Big Data 
engines.

@@

Apache Spark is a popular Scala-based processing framework for Big Data, with
APIs in Java, Python and R. Its 
generalized nature allows Spark to not only be applied to batch workflows,
but also SQL queries, iterative machine learning applications, 
data streaming and graph processing. Spark's
main features include data-locality, in-memory processing and lazy evaluation,
which it achieves through its principle abstraction, the Resilient Distributed 
Dataset (RDD).

An RDD is an immutable parallel data structure that achieve fault-tolerance 
through the concept of lineage~\cite{RDDpaper}. Rather than permitting
fine-grained transformations, only coarse-grained transformations (applying to
many elements in the RDD), can be applied, thereby making it simple to keep a 
log of how the data was modified. This log, known as the lineage, is used
to reproduce any lost data.

Two types of operations can be performed on RDDs:
transformations and actions. Applying a transformation to an RDD produces a new,
child, RDD through a narrow or wide dependency. A narrow dependency signifies 
that the child is only dependent on a single parent partition, whereas a child 
RDD is dependent on all parent partitions in a wide dependency. Examples of 
spark transformations include map, filter and join. To materialize an RDD, an
action must be performed, such as a reduce or a collect.

All actions and wide dependencies require a shuffle -- Spark's most costly
operation. Every shuffle begins with each map task saving its data to local
files to be fetched by the shuffle operation. The shuffle operation then 
redistributes the data across the partitions as requested. A shuffle marks a 
stage boundary in Spark. In other words, a shuffle will not begin until all 
dependent map tasks have completed.

Spark will spill RDD data to disk whenever the complete dataset does not fit in
memory. Moreover, as Spark transformations generate new RDDs and numerous 
transformations may occur in a single application, Spark implements a LRU eviction
policy. If an evicted RDD needs to be reused, Spark will recomputed it using the
lineage data collected. It therefore becomes of great importance to persist to
memory and/or 
disk RDDs which will be reused throughout the application, particularly if the
recomputation is costly. 
%mention results in thunder paper

Data locality in Spark is achieved through the  scheduling task to partitions which have the
data loaded in memory. If the data is instead stored on HDFS, the scheduler will
assign it to one of the preferred locations specified by HDFS. Spark's scheduler
utilizes delay scheduling to ensure optimal fairness and locality for all tasks.
%improve paragraph

Three different types of schedulers are compatible with Spark: 1) Spark 
Standalone, 2) YARN~\cite{yarn} and 3) Mesos~\cite{mesos}. The Spark Standalone
scheduler is the default scheduler. YARN was a scheduler designed for Hadoop~\cite{hadoop}
clusters. It is prepackaged with Hadoop installations. In contrast to YARN, 
Mesos was designed to be used in multitenant cluster environments. In our experiments,
we focus on the Spark Standalone cluster.

Executing Spark applications on HPC systems with Spark-unaware schedulers may
be inefficient. The amount of resources requested by Spark may impede Spark-cluster
scheduling time. Using pilot-scheduling strategies to add nodes to the Spark cluster
as they are allocated by the underlying schedulers may speedup allocation and 
overall processing time. This, however, is not studied in the current paper.

As Spark is frequently used by the scientific community, we designed our 
experiments using the PySpark API. This is at a cost to performance as the PySpark
code must undergo Python to Java serialization.
